Iterative Monomorphisation

### 2024-02-19 iterative_monomorphisation.tex
first proper draft for preliminaries

% TODO, at the end check the following
%   - formulas -> formulae
%   - clause -> formula
%   - monomorphization -> monomorphisation
%   - higer order -> higer-order
%   - TH1 with ugly 1
%   - correctly spelled TPTP, Leo-III, Zipperposition, E, Vampire
%   - use of formula vs clause
%   - polymorphic and monomorphic
%   - title of bits of the algorithms
%   - eg or ex -> e.g.

@@

{\obeylines

\noindent\kern\parindent
  * algorithm works a priori with any rank-1 polymorphic logic

  * in particular, it works with polymorphic first- and higher-order logic as 
    embodied by TPTP's TF1 and TH1 syntaxes \cite{blanchette-paskevich-201x,th1}

  * the algorithm does not need to inspect the precise structure of formulae
  * instead, it works on monomorphic or polymorphic instances of symbols that 
    occur in the formulae
  * type quantification is implicit, meaning that all type variables occurring
    in a formula are assumed to be universally quantified at the formula's top
    level
}

@@

\begin{definition}%[Function symbol]
A (function or predicate) \emph{symbol} \(f\) has a type arity that defines the number of type arguments it takes. These arguments are listed between angle brackets: \(f\langle \tau_1, \dots, \tau_n\rangle\), where each $\tau_i$ is a type.
\end{definition}

{
\obeylines

* substitutions

* matching
}

>>>>

% TODO, at the end check the following
%   - formulas -> formulae
%   - clause -> formula
%   - monomorphization -> monomorphisation
%   - higer order -> higer-order
%   - TH1 with ugly 1
%   - correctly spelled TPTP, Leo-III, Zipperposition, E, Vampire
%   - use of formula vs clause
%   - polymorphic and monomorphic
%   - title of bits of the algorithms
%   - eg or ex -> e.g.
%   - function symbol -> symbol instance (depends on context)
%   - ... -> \dots
%   - comma splices!!!!
%   - ``'' -> `'
%   - use \Phi instead of F for formulae
%   - concrete types -> monomorphic types
%   - use in not \in when we are not precisely dealing with set membership
%
%

@@

This algorithm does not need to inspect the structure of the problem's formulae because it relies exclusively on the formulae's monomorphic and polymorphic symbol instances. The notion of formula is necessary as type variables are assumed to be universally quantified at a formula's top level. The precise form of a formula can nevertheless be left unspecified.

Iterative monomorphisation can therefore be used in a vast range of rank-1 polymorphic logic. In particular, it can operate in the polymorphic first- and higher-order logics embodied by TPTP's TF1 and TH1 syntaxes \cite{blanchette-paskevich-2013,th1}, popular among automatic provers.

The following terms are used in the algorithm descriptions.

%{\obeylines
%
%\noindent\kern\parindent
%  * algorithm works a priori with any rank-1 polymorphic logic
%
%  * in particular, it works with polymorphic first- and higher-order logic as 
%    embodied by TPTP's TF1 and TH1 syntaxes \cite{blanchette-paskevich-2013,th1}
%
%  * the algorithm does not need to inspect the precise structure of formulae
%  * instead, it works on monomorphic or polymorphic instances of symbols that 
%    occur in the formulae
%  * type quantification is implicit, meaning that all type variables occurring
%    in a formula are assumed to be universally quantified at the formula's top
%    level
%}

@@

\begin{definition}%[Function symbol]
A (function or predicate) \emph{symbol} \(f\) has a type arity that defines the number of type arguments it takes. These arguments are listed between angle brackets: \(f\langle \tau_1, \dots, \tau_n\rangle\), where each $\tau_i$ is a type.

A \emph{symbol instance} refers to a symbol provided with its type arguments.
\end{definition}

\begin{definition} % HEAVILY inspired from 2.4 in mono-trans (it is the definition of a substitution, there's bound to be some comonality between all of them) 
A \emph{substitution} \(\sigma\) is a function mapping one or more type variables to corresponding monomorphic type. Substitutions are in practice mostly be applied to formulae. Given a formula \(\Phi\), the formula obtained by replacing each type variable \(\alpha\) in \(\Phi\) by \(\sigma\alpha\) is written \(\sigma\Phi\).

\end{definition}

\begin{definition}
   Given a monomorphic type \(\tau_1\) and a polymorphic type \(\tau_2\), \emph{matching} \(\tau_2\) against \(\tau_1\) will either fail or yield a substitution \(\sigma\) such that \(\sigma\tau_2 = \tau_1\). Note that in the second case \(\sigma\tau_2\) is monomorphic.
\end{definition}

%{
%\obeylines
%
%* substitutions
%
%* matching
%}
